---
title: "Tutorial 1: predict the agent"
author: "Chixiang Chen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 1: predict the agent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/chixiang.chen/OneDrive - University of Maryland School of Medicine/phd_in_psu/research/Prof._Wu/Network/tissue/blood_vessel/')
```

**DRDNetPro** is a bio-protocol for recovering disease risk-associated pseudo-dynamic networks (DRDNet) from steady-state data. It incorporates risk prediction model of having certain disease, a varying coefficient model, multiple ordinary differential equations, and group lasso estimation to learn a series of networks. This tutorial will provide detailed information for the first step of implementing the [DRDNet](https://academic.oup.com/bioinformatics/article-abstract/38/9/2481/6537533). The data we will use in this tutorial is named [pheno_used](https://github.com/chencxxy28/DRDNetPro/raw/main/vignettes/data/pheno_used.csv).

To begin with, we need to install the following packages in R

# Installation
```{r, eval=FALSE}
if (!require("devtools")) {
  install.packages("devtools")
}
devtools::install_github("chencxxy28/DRDNetPro")
```


```{r,message=FALSE}
library(pROC)
library(ranger)
library(xgboost)
library(dplyr)    

readRDSFromWeb <- function(ref) {
  readRDS(gzcon(url(ref)))
}
```

# Prepare the data for training the agent
The agent in the following tutorial is related to the risk of having hypertension. Before predicting the agent, the users need to prepare the data for training the model. The data is required to be a matrix, which includes the outcome in the first column and all potential predictors in the remaining columns. Do not include the intercept as one column. The outcome in this protocol is in the binary scale (e.g., 0=no disease and 1=disease), which in practice indicates the disease status. All values in the matrix should be numeric. The missing data is allowed in this training process, though it is always recommended to have a complete data as an input. The following working data is from one sample of data in GTEx project and de-identified. 

```{r,message=FALSE}
#read the data
pheno_raw<-read.csv("https://github.com/chencxxy28/DRDNetPro/raw/main/vignettes/data/pheno_did.csv")

# pheno_raw<-read.csv("https://github.com/chencxxy28/DRDNetPro/raw/main/vignettes/data/pheno_used.csv")

# pheno_raw$SUBJID<-rep(1:nrow(pheno_raw))
# pheno_raw<-pheno_raw[,-which(colnames(pheno_raw)=="MHSMKSTS")]
# colnames(pheno_raw)<-c("ID","V1","V2","V3","V4","V5","V6",
#                        "V7","V8","V9","disease","V10",
#                        "V11")
# write.csv(pheno_raw,"pheno_did.csv",row.names = FALSE)

#rearrange the data
pheno_used<-data.frame(disease=pheno_raw$disease,pheno_raw[,!(colnames(pheno_raw) %in% c("disease","ID"))])
head(pheno_used)
```

# Build up the training model
There is no unique way to predict the agent (risk of having hypertension). Statistical regression models or machine learning algorithms can be applied. In our program, we allow the users to specify their own method, which includes conventional logistic regression, random forest, and XGBoost. The final imputed agent is supported between 0 and 1, which quantifies the disease risk of interest.

## logistic regression
```{r,message=FALSE}
fit<-glm(disease~.,data=pheno_used,family = binomial(link = "logit"))
rate<-fitted(fit)
summary(rate)
```

## Random forest
Random forest is an ensemble of decision trees. It builds and combines multiple decision trees to get more accurate predictions. It’s a non-linear classification algorithm. Each decision tree model is used when employed on its own. An error estimate of cases is made that is not used when constructing the tree. This is called an out of bag error estimate mentioned as a percentage. There are several R packages available: one is `randomForest`, and one is `ranger`. Please refer to [random forest](http://www.css.cornell.edu/faculty/dgr2/_static/files/R_html/CompareRandomForestPackages.html#2_Random_forest_with_randomForest) for more details. This section provides a demo code based on the package `ranger`.

Create the data for `ranger`, the outcome should be a factor.
```{r,message=FALSE}
mydata<-pheno_used
mydata$MHHTN<-as.factor(mydata$MHHTN)
```

Run the following code to train the model. Note that some parameters, such as the number of variables, can be further tuned. Please refer [random forest](http://www.css.cornell.edu/faculty/dgr2/_static/files/R_html/CompareRandomForestPackages.html#2_Random_forest_with_randomForest) for more information.
```{r,message=FALSE}
set.seed(12345)
rf<-ranger(MHHTN~., num.trees=1000,data=mydata,replace=T,write.forest = T,probability = T)
rate.rf<-predict(rf, mydata)$predictions[,2]
summary(rate.rf)
```

## XGBoost
XGBoost an efficient implementation of gradient boosting framework. It provides built-in k-fold cross-validation
Stochastic GBM with column and row sampling (per split and per tree) for better generalization,
includes efficient linear model solver and tree learning algorithms, and supports various objective functions, including regression, classification and ranking. Please refer to [gradient boosting](http://uc-r.github.io/gbm_regression) for more details. This section provides a demo code based on the package `XGBoost`.

XGBoost requires a specific type of input data, which can be run by the following code

```{r,message=FALSE}
mydata<-pheno_used

# variable names
features <- setdiff(names(mydata), "MHHTN")
# Create the treatment plan from the training data
treatplan <- vtreat::designTreatmentsZ(mydata, features, verbose = FALSE)
# Get the "clean" variable names from the scoreFrame
new_vars <- treatplan %>%
  magrittr::use_series(scoreFrame) %>%        
  dplyr::filter(code %in% c("clean", "lev")) %>% 
  magrittr::use_series(varName) 
# Prepare the training data
features_train <- vtreat::prepare(treatplan, mydata, varRestriction = new_vars) %>% as.matrix()
response_train <- mydata$MHHTN
```

Tune the number of trees that minimizes the error
```{r,message=FALSE}
set.seed(12345)
xgb.fit1 <- xgb.cv(
  data = features_train,
  label = response_train,
  nrounds = 1000,
  nfold = 5,
  objective = "binary:logistic",  # for regression models
  verbose = 0 ,              # silent,
  early_stopping_rounds = 10 # stop if no improvement for 10 consecutive trees
)

vect=xgb.fit1$evaluation_log %>%
  dplyr::summarise(
    ntrees.train = which(train_logloss_mean == min(train_logloss_mean))[1],
    rmse.train   = min(train_logloss_mean),
    ntrees.test  = which(test_logloss_mean == min(test_logloss_mean))[1],
    rmse.test   = min(test_logloss_mean),
  )
```

Run the final trained model to obtain the predicted agent
```{r,message=FALSE}
set.seed(12345)
#final estimate
xgb.fit.final <- xgboost(
  data = features_train,
  label = response_train,
  nrounds = as.numeric(vect[3]),
  objective = "binary:logistic",  # for regression models
  verbose = 0              # silent,
)
rate.xgb = predict(xgb.fit.final, features_train,type="response")
summary(rate.xgb)
```

# Predict and evaluate the agent
When the agent is imputed, it is better to check its prediction performance. Note that the pseudo-dynamic network is sensitive to the values of agent, and the agent with good prediction to the disease risk is more preferred. To check its performance, we can use ROC and AUC value. For evaluation purpose, we need to create a training data and testing data. The training data is used for data fitting, whereas the testing data is used for prediction evaluation. Below, we consider 90% of data for training and 10% of data for testing.
```{r,message=FALSE}
set.seed(12345)
sample.ind<-sample(1:nrow(pheno_used),round(0.9*nrow(pheno_used))) #id used for training
pheno_used_training<-pheno_used[sample.ind,]
pheno_used_testing<-pheno_used[!(1:nrow(pheno_used) %in% sample.ind),]

fit_train<-glm(disease~.,data=pheno_used_training,family = binomial(link = "logit"))

test_prob = predict(fit_train, newdata = pheno_used_testing, type = "response")
test_roc_specific =roc(pheno_used_testing$disease ~ test_prob, plot = TRUE, print.auc = TRUE)
```

The AUC value is around 0.83, which is desired. Similar procedure could be applied to the case with random forest or XGBoost. After the agent imputation, we can move forward to the stage of network learning, which is illustrated in [Tutorial 2](https://chencxxy28.github.io/DRDNetPro/articles/web/Tutorial%202.html). Before the end of this tutorial, let us create a phenotypical data consisting of the imputed agent for the use in next tutorial.

```{r,message=FALSE}
pheno_subject_id<-substring(pheno_raw$SUBJID,6,10)
pheno_together<-data.frame(cbind(pheno_subject_id,rate,pheno_used$MHSMKSTS,pheno_used$SEX,pheno_used$HGHT,pheno_used$BMI,pheno_used$MHHTN))
colnames(pheno_together)<-c("id","rate","smoke","sex","height","bmi","hyper")
head(pheno_together)
write.csv(pheno_together,"pheno_together.csv",row.names =F)
```

This is the end of the Tutorial 1, please visit [Tutorial 2](https://chencxxy28.github.io/DRDNetPro/articles/web/Tutorial2.html) for the next step of network learning 









